---
title: |
  External Validation of Berten's Model for Predicting COPD Exacerbations
type: ARTICLE TEMPLATE
author:
  - name: Amin Adibi
    affil: a
    email: amin.adibi@ubc.ca
  - name: Don Sin
    affil: b
    email: don.sin@hli.ubc.ca
  - name: Mohsen Sadatsafavi
    affil: a, \dagger, \ddagger
    email: mohsen.sadatsafavi@ubc.ca
affiliation:
  - num: a
    address: |
      Respiratory Evaluation Sciences Program, Faculty of Pharmaceutical Sciences, University of British Columbia, Vancouver, BC, Canada
  - num: b
    address: |
      Division of Respiratory Medicine, Department of Medicine, The UBC Centre for Heart Lung Innovation, St. Paul's Hospital, University of British Columbia, Vancouver, BC, Canada
bibliography: RESP.bib
appendix: appendix.tex
abstract: |
  We are reporting an independent validation of the COPD exacerbation prediction model
  developed by Bertens and colleagues in the 
keywords: |
  COPD; prediction; validation
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

# Introduction

The Berten's model [@bertens_development_2013].



# Methods

We have followed recommendations of the Transparent Reporting of a Multivariable Prediction Model for Individual Prognosis or Diagnosis (the TRIPOD Statement) [@collins_transparent_2015] in reporting our independent validation of the Bertens model. External validation was done using three years of data from COPD patients enrolled in the Evaluation of COPD Longitudinally to Identify Predictive Surrogate End-points (the ECLIPSE Study), an independent non-interventional observational longitudinal COPD cohort study that was aimed at characterizing progression of the disease[@agusti_characterisation_2010].

The first year of the ECLIPSE study was used to establish exacerbation history, year two was used to assess validity of one-year and two-year predictions were assessed using data from year two, and years 2-3, respectively. 

We report external validation of the model in terms of calibration (the degree to which predicted and observed risk of exacerbations are in agreement), discrimination (the extend to which the model is able to separate higher and lower risk individuals), and clinical utility (net benefit across different risk thresholds for treatment decision). 

Calibration was assessed through visual examination of the calibration curve. Differentiation was assessed by plotting the Receiver Operating Characteristic (ROC) curve and its area-under-the-curve (AUC), also known as the c-statistic. Clinical utility was assesses through have reported Decision Curve Analysis [@vickers_simple_2019]. 


The study was approved by the University of British Columbia and Providence Health Research Ethics Board (H11–00786).
<!-- If your manuscript has supplementary content you can prepare this using the `suppldata` document-class option, which will suppress the `article history' date. This option must _not_ be used on any primary content. -->

# Results


# Discsussions


## Code chunks

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r dataprep, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(bertens)
library(pROC)
library(gbm)
library(rmda)
library(ggthemes)
library(naniar)
eclipse.raw <- read_csv("../eclipse.csv") %>% 
  filter (VISIT == "1 Year") %>% 
  select (id = SUBJECT_ID, Age = AGE, sex = SEX, 
          fev1 = FEV1PSPC) %>%
          mutate(id = str_remove(id, "xyl"))

# better cardiovascular data also available.

packyear <- read_table2("../packyear.txt") %>% select (id = SUBJECT_ID, packyears = SUPKYR) %>%
                                            mutate(id = str_remove(id, "ecl"))

cv_cond <- read_csv("../cv_cond.csv") %>% select (id= SUBJECT_ID, 
                                               strokeHx = ATS8E,
                                               heartAttackHx = ATS8D) %>%
                                       mutate(id = str_remove(id, "xyl"))


exacerbation <- read_csv("../exacerbation.csv") %>% select(id= SUBJECT_ID, 
                                                        year = YRCAT,
                                                        modSevereExac = EXACNUM1,
                                                        sevExac = SEVEXN1) %>%
                                                        #followup = OBSTIME) %>%
               pivot_wider(names_from = year, values_from = c(modSevereExac, sevExac)) %>%
               mutate(year1 = `modSevereExac_Year 1`,
                      year2 = `modSevereExac_Year 2`,
                      year2to3 = `modSevereExac_Years 1-3` - year1,
                      year1Sev = `sevExac_Year 1`,
                      id = str_remove(id, "xyl")) %>%
               select (id, year1, year1Sev, year2, year2to3)


# the paper defines A history of vascular disease was defined as stroke, transient ischemic attack, or peripheral arterial disease.
# we used strokeHx | heartAttackHx as a proxy for it in ECLIPSE.
eclipse <- eclipse.raw %>% filter (fev1<85) %>%  # removing non-copd controls
                           left_join(cv_cond, by = "id") %>% left_join(packyear, by = "id") %>%
                           mutate(strokeHx  = recode (strokeHx,
                                                      "Y" = "TRUE",
                                                      "N" = "FALSE",
                                                      "U" = "NA")) %>%
                           mutate(heartAttackHx  = recode (heartAttackHx,
                                                      "Y" = "TRUE",
                                                      "N" = "FALSE")) %>%
                           mutate(heartAttackHx = as.logical(heartAttackHx),
                                  strokeHx = as.logical(strokeHx),
                                  vascularDx = (strokeHx | heartAttackHx)) %>%
                          select(-strokeHx, -heartAttackHx) %>%
                          left_join(exacerbation, by = "id") %>%
                        #  filter(packyears > 0) %>%
                          mutate(Observed_Exac_in2to3 = ifelse(year2to3>0, 1, 0))  %>%
                          mutate(Observed_Exac_in2 = ifelse(year2>0, 1, 0))  %>%
                          mutate(year1FreqExacerbator = ifelse(((year1>1) | (year1Sev>0)), 1, 0),
                                 year1HadExac = ifelse(year1>0, 1, 0),
                                 predictedBertens = 
                                   bertens(exacerbationHx = year1HadExac, 
                                           fev1=fev1, 
                                           packYears = packyears,
                                           vascularDx = vascularDx),
                                 predictedBertens1Yr = 1-sqrt(1-predictedBertens)) 

eclipseComplete <- eclipse %>% drop_na()

eclipseCompleteSmokers <- eclipseComplete %>% filter (packyears>=1)

write.csv(eclipse, "eclipse_bertens.csv")

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE, out.width='80%'}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Some guidelines for using the standard features of \LaTeX

## Sections

The \textsf{Interact} layout style allows for five levels of section heading, all of which are provided in the `interact` class file using the standard \LaTeX\ commands `\section`, `\subsection`, `\subsubsection`, `\paragraph` and `\subparagraph`. Numbering will be automatically generated for all these headings by default.

## Lists

Numbered lists are produced using the `enumerate` environment, which will number each list item with arabic numerals by default. For example,

1. first item
2. second item
3. third item

Alternative numbering styles can be achieved by inserting an optional argument in square brackets to each `item`, e.g.\ `\item[(i)] first item`\, to create a list numbered with roman numerals at level one.

Bulleted lists are produced using the `itemize` environment. For example,

* First bulleted item
* Second bulleted item
* Third bulleted item

## Figures

```{r pressure-plot, fig.show='hide'}
plot(pressure)
```

The `interact` class file will deal with positioning your figures in the same way as standard \LaTeX. It should not normally be necessary to use the optional `[htb]` location specifiers of the `figure` environment in your manuscript; you may, however, find the `[p]` placement option or the `endfloat` package useful if a journal insists on the need to separate figures from the text.

Figure captions appear below the figures themselves, therefore the `\caption` command should appear after the body of the figure. For example, Figure~\ref{sample-figure} with caption and sub-captions is produced using the following commands:
```
\begin{figure}
\centering
\subfigure[An example of an individual figure sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{path/to/fig}}}\hspace{5pt}
\subfigure[A slightly shorter sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{path/to/fig}}}
\caption{Example of a two-part figure with individual sub-captions
 showing that captions are flush left and justified if greater
 than one line of text.} \label{sample-figure}
\end{figure}
```
\begin{figure}
\centering
\subfigure[An example of an individual figure sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{`r knitr::fig_chunk('pressure-plot', 'pdf')`}}}\hspace{5pt}
\subfigure[A slightly shorter sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{`r knitr::fig_chunk('pressure-plot', 'pdf')`}}}
\caption{Example of a two-part figure with individual sub-captions
 showing that captions are flush left and justified if greater
 than one line of text.} \label{sample-figure}
\end{figure}

To ensure that figures are correctly numbered automatically, the `\label` command should be included just after the `\caption` command, or in its argument.

The `\subfigure` command requires `subfigure.sty`, which is called in the preamble of the `interacttfssample.tex` file (to allow your choice of an alternative package if preferred) and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Please supply any additional figure macros used with your article in the preamble of your .tex file.

The source files of any figures will be required when the final, revised version of a manuscript is submitted. Authors should ensure that these are suitable (in terms of lettering size, etc.) for the reductions they envisage.

The `epstopdf` package can be used to incorporate encapsulated PostScript (.eps) illustrations when using PDF\LaTeX, etc. Please provide the original .eps source files rather than the generated PDF images of those illustrations for production purposes.


## Tables

The `interact` class file will deal with positioning your tables in the same way as standard \LaTeX. It should not normally be necessary to use the optional `[htb]` location specifiers of the `table` environment in your manuscript; you may, however, find the `[p]` placement option or the `endfloat` package useful if a journal insists on the need to separate tables from the text.

The `tabular` environment can be used as shown to create tables with single horizontal rules at the head, foot and elsewhere as appropriate. The captions appear above the tables in the \textsf{Interact} style, therefore the `\tbl` command should be used before the body of the table. For example, Table~\ref{sample-table} is produced using the following commands:
\begin{table}
\tbl{Example of a table showing that its caption is as wide as
 the table itself and justified.}
{\begin{tabular}{lcccccc} \toprule
 & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7}
 Class & One & Two & Three & Four & Five & Six \\ \midrule
 Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\
 Beta & B2 & B2 & B3 & B4 & B5 & B6 \\
 Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule
\end{tabular}}
\tabnote{\textsuperscript{a}This footnote shows how to include
 footnotes to a table if required.}
\label{sample-table}
\end{table}
```
\begin{table}
\tbl{Example of a table showing that its caption is as wide as
 the table itself and justified.}
{\begin{tabular}{lcccccc} \toprule
 & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7}
 Class & One & Two & Three & Four & Five & Six \\ \midrule
 Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\
 Beta & B2 & B2 & B3 & B4 & B5 & B6 \\
 Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule
\end{tabular}}
\tabnote{\textsuperscript{a}This footnote shows how to include
 footnotes to a table if required.}
\label{sample-table}
\end{table}
```

To ensure that tables are correctly numbered automatically, the `\label` command should be included just before `\end{table}`.

The `\toprule`, `\midrule`, `\bottomrule` and `\cmidrule` commands are those used by `booktabs.sty`, which is called by the `interact` class file and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Tables produced using the standard commands of the `tabular` environment are also compatible with the `interact` class file.


## Landscape pages

If a figure or table is too wide to fit the page it will need to be rotated, along with its caption, through 90$^{\circ}$ anticlockwise. Landscape figures and tables can be produced using the `rotating` package, which is called by the `interact` class file. The following commands (for example) can be used to produce such pages.
```
\setcounter{figure}{1}
\begin{sidewaysfigure}
\centerline{\epsfbox{figname.eps}}
\caption{Example landscape figure caption.}
\label{landfig}
\end{sidewaysfigure}
```
```
\setcounter{table}{1}
\begin{sidewaystable}
 \tbl{Example landscape table caption.}
  {\begin{tabular}{@{}llllcll}
    .
    .
    .
  \end{tabular}}\label{landtab}
\end{sidewaystable}
```
Before any such float environment, use the `\setcounter` command as above to fix the numbering of the caption (the value of the counter being the number given to the preceding figure or table). Subsequent captions will then be automatically renumbered accordingly. The `\epsfbox` command requires `epsfig.sty`, which is called by the `interact` class file and is also included in the \textsf{Interact} \LaTeX\ bundle for convenience.

Please note that if the `endfloat` package is used, one or both of the commands
```
\DeclareDelayedFloatFlavor{sidewaysfigure}{figure}
\DeclareDelayedFloatFlavor{sidewaystable}{table}
```
will need to be included in the preamble of your .tex file, after the `endfloat` package is loaded, in order to process any landscape figures and/or tables correctly.


# Acknowledgement(s) {-}

An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Disclosure statement {-}

An unnumbered section, e.g.\ `\section*{Disclosure statement}`, may be used to declare any potential conflict of interest and included _in the non-anonymous version_ before any Notes or References, after any Acknowledgements and before any Funding information.

# Funding {-}

An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Notes on contributor(s) {-}

An unnumbered section, e.g.\ `\section*{Notes on contributors}`, may be included _in the non-anonymous version_ if required. A photograph may be added if requested.

# Nomenclature/Notation {-}

An unnumbered section, e.g.\ `\section*{Nomenclature}` (or `\section*{Notation}`), may be included if required, before any Notes or References.

# Notes {-}

An unnumbered `Notes` section may be included before the References (if using the `endnotes` package, use the command `\theendnotes` where the notes are to appear, instead of creating a `\section*`).

# References

## References cited in the text

## The list of references

References should be listed at the end of the main text in alphabetical order by authors' surnames, then chronologically (earliest first). If references have the same author(s), editor(s), etc., arrange by year of publication, with undated works at the end.
A single-author entry precedes a multi-author entry that begins with the same name.
If the reference list contains two or more items by the same author(s) in the same year, add a, b, etc. and list them alphabetically by title.
Successive entries by two or more authors when only the first author is the same are alphabetized by co-authors' surnames.
If a reference has more than ten named authors, list only the first seven, followed by 'et al.'.
If a reference has no author or editor, order by title; if a date of publication is impossible to find, use `n.d.' in its place.

The following list shows some sample references prepared in the Taylor & Francis Chicago author-date style.

[@Ade09; @Alb05]
